---
title: "Calculating_MOB_interaction_networks"
author: "Roey Angel"
date: "13 October 2015"
output: pdf_document
---

# Network analysis for the dataset derived from Zheng et al 2014. Biogeosciences.
```{r general}
library(dplyr)
library(igraph)
library(vegan)
library(ggplot2)
library(grid)
library(scales)
library(RColorBrewer)
library(my.tools)
setwd("~/Networks/Zheng")
theme_set(theme_gray(base_size = 18, base_family = "sans"))
```

```{r functions}
PrepOtuTax <- function(Taxonomy, tax.level = "Class", ...){
  ## turn a .taxa file into a data frame
  Taxonomy <- data.frame(row.names = rownames(Taxonomy), Size = Taxonomy$Size, do.call("rbind", strsplit(Taxonomy$Taxonomy, ";", fixed = TRUE))) # separate taxa levels to column with ;

  Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("\\(.*\\)", "", x))) # remove bs values (irrelevant)
  Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("\"","",x))) # remove "
  Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("_.*$","",x))) # remove _
  Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("uncultured", "Unclassified", x)))
  Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("unclassified", "Unclassified", x)))
  Taxonomy$Size <- as.numeric(levels(Taxonomy$Size)[Taxonomy$Size])
  colnames(Taxonomy) <- c("Size","Domain","Phylum","Class","Order","Family","Genus")
  
  
  group_by_(Taxonomy, tax.level) %>% dplyr::summarise(Rel.abun = sum(Size)/sum(Taxonomy$Size)*100) -> Tax.sizes

  Rares <- unlist(lapply(Tax.sizes[Tax.sizes$Rel.abun <= 1, tax.level], as.character), use.names = FALSE)
  
  Taxonomy[, tax.level] <- factor(Taxonomy[, tax.level], levels = c(levels(Taxonomy[, tax.level]), "Rare")) # add Rare level
  Taxonomy[which(as.character(lapply(Taxonomy[, tax.level], as.character)) %in% Rares), tax.level] <- "Rare"
  
  return(Taxonomy)
}

PrepNetDF <- function(graph, layout_function = layout.fruchterman.reingold, membership = FALSE) {
  # Generate layout
  set.seed(170922)
  l <- layout_function(graph)

  # Get groups to colour
  # colour <- get.vertex.attribute(graph,'colour')
  colour <- graph$colour
  # Get node names
  name <- get.vertex.attribute(graph, 'name')
  # Get OTU abundances
  # abundance <- get.vertex.attribute(graph,'abundance')
  abundance <- graph$abundance
  # Get edge weights
  weight <- get.edge.attribute(graph, 'weight')
  if (is.null(weight)) weight <- rep(1, length(get.edgelist(graph))/2)
  direction <- factor(weight / abs(weight), levels = c(1,-1), labels = c("Positive", "Negative"))
  
  # Make data frame for ggplot
  coords <- data.frame(cbind(name, colour), abundance, l, stringsAsFactors = T)
  coords$colour <- factor(coords$colour, labels = levels(as.factor(colour)))

  # Add edges
  edges <- data.frame(get.edgelist(graph), abs(weight), direction)
  # edges <- edges[, ]
  edges <- do.call(rbind, apply(edges, 1, function(e) data.frame(coords[coords$name == e[1], ], coords[coords$name == e[2], ], weight = e[3], direction = e[4])))
  edges$weight <- as.numeric(levels(edges$weight)[edges$weight])
  edges$midX  <- (edges$X1 + edges$X1.1) / 2
  edges$midY  <- (edges$X2 + edges$X2.1) / 2

  if (membership) { # network depicting membership?
    coords$methanotroph <- graph$methanotroph
  }

  return(list(coords, edges))
}

blankground <- function() {
  theme(panel.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.margin = unit(0,"null"),
        plot.margin = rep(unit(0,"null"),4),
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_blank()
  )
}

GGNetwork <- function(coords, edges, tax.level = "Class", OTU.labels = FALSE) {
  ## Plot OTU network
  blankground <- function() {
    theme(panel.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.margin = unit(0,"null"),
          plot.margin = rep(unit(0,"null"),4),
          axis.ticks = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank()
    )
  }

  p <- ggplot()  +
    geom_segment(data = edges, aes(x = X1, y = X2, xend = X1.1, yend = X2.1, size = weight, colour = direction), alpha = 1/10) + # , colour = "#55555533"
    geom_point(data = coords, aes(X1, X2, fill = colour), shape = 21, size = 5, alpha = 1/2) +
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL) +
    blankground() +
    guides(colour = guide_legend(title = "Direction", override.aes = list(alpha = 1/2, size = 4)), fill = guide_legend(title = tax.level), size = guide_legend(title = "Correlation strength")) +
    scale_size(range = c(2, 10))
  #     scale_colour_brewer(palette="Dark2")

  if (OTU.labels) {
    p + geom_text(aes(x = X1, y = X2, label = name), data = coords[coords$abundance > 100, ], alpha = 1/2) #colour="#707070"
    #   geom_text(aes(x= X1, y= X2, label = name), data = coords[coords$abundance>100, ], position=position_dodge(0.9), alpha = 1/2) + #colour="#707070"
    #   geom_text(aes(midX, midY,label=weight,size=10), data=edges) +
  } else p
}

GGNetworkPoints <- function(coords, tax.level = "Class", OTU.labels = FALSE) {
  p <- ggplot()  +
    geom_point(data = coords, aes(X1, X2, colour = colour, size = abundance, shape = methanotroph), alpha = 4/5) +
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL) +
    blankground() +
    guides(colour = guide_legend(title = tax.level, override.aes = list(alpha = 1/2, size = 4)), fill = guide_legend(title = tax.level), size = guide_legend(title = "Rel. abundance (%)"), shape = guide_legend(title = "Methanotroph")) +
    scale_size(range = c(2, 10)) + 
    scale_colour_manual(values = brewer24) 
  
  if (OTU.labels) {
    p + geom_text(aes(x = X1, y = X2, label = gsub("OTU_([0-9]+)", "\\1", name)), data = coords[coords$abundance > 1, ], alpha = 4/5) #colour="#707070"
    #   geom_text(aes(x= X1, y= X2, label = name), data = coords[coords$abundance>100, ], position=position_dodge(0.9), alpha = 1/2) + #colour="#707070"
    #   geom_text(aes(midX, midY,label=weight,size=10), data=edges) +
  } else p
}

GGNetworkSegments <- function(edges) {
  p <- ggplot()  +
    geom_segment(data = edges, aes(x = X1, y = X2, xend = X1.1, yend = X2.1, size = weight, colour = direction), alpha = 1/20) + # , colour = "#55555533"
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL) +
    blankground() +
    guides(colour = guide_legend(title = "Direction", override.aes = list(alpha = 1/2, size = 4)), size = guide_legend(title = "Correlation strength")) +
    scale_size(range = c(2, 10)) + 
    scale_colour_manual(values = c("#377EB8", "#E41A1C"))
}

GGNetworkMethano <- function(coords, edges, tax.level = "Class", OTU.labels = FALSE) {
  ## Plot OTU network
  blankground <- function() {
    theme(panel.background = element_blank(),
          panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.margin = unit(0,"null"),
          plot.margin = rep(unit(0,"null"),4),
          axis.ticks = element_blank(),
          axis.text.x = element_blank(),
          axis.text.y = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank()
    )
  }

  p <- ggplot()  +
    geom_segment(data = edges, aes(x = X1, y = X2, xend = X1.1, yend = X2.1, size = weight), colour = "#377EB8", alpha = 1/15) + # , colour = "#55555533"
    geom_point(data = coords, aes(X1, X2, colour = colour, shape = methanotroph), size = 5, alpha = 1/2) +
    scale_x_continuous(breaks = NULL) + scale_y_continuous(breaks = NULL) +
    blankground() +
    guides(colour = guide_legend(title = "Direction", override.aes = list(alpha = 1/2, size = 4)), fill = guide_legend(title = tax.level), size = guide_legend(title = "Correlation strength")) +
    scale_size(range = c(2, 10))
  #     scale_colour_brewer(palette="Dark2")

  if (OTU.labels) {
    p + geom_text(aes(x = X1, y = X2, label = name), data = coords[coords$abundance > 100, ], alpha = 1/2) #colour="#707070"
    #   geom_text(aes(x= X1, y= X2, label = name), data = coords[coords$abundance>100, ], position=position_dodge(0.9), alpha = 1/2) + #colour="#707070"
    #   geom_text(aes(midX, midY,label=weight,size=10), data=edges) +
  } else p
}

brewer24 <- c(RColorBrewer::brewer.pal(n = 8, "Set1"), RColorBrewer::brewer.pal(n = 8, "Dark2"), RColorBrewer::brewer.pal(n = 7, "Accent"), '#525252')
```

## Calculate correlations
### Prepare data 

### Trim data, remove OTUs with less than 10 reads in total and which only appear < 20% of the samples
```{r, cache=TRUE}
Zheng_OTU <- read.table("otuCountTable_13C-CH4_SIP.txt", header = TRUE, row.names = 1)
Zheng_OTU <- Zheng_OTU[order(rowSums(Zheng_OTU), decreasing = TRUE), ]
Zheng_OTU <- Zheng_OTU[rowSums(Zheng_OTU) > 10, ]
Zheng_OTU <- Zheng_OTU[rowSums(apply(Zheng_OTU, 2, function(x) x > 0)) > (ncol(Zheng_OTU) * 0.2), ]
# are there samples with no OTUs?
(drop <- which(colSums(Zheng_OTU) == 0))
if (length(drop) > 0) Zheng_OTU <- Zheng_OTU[, -drop]
(1 - mean(vegdist(t(Zheng_OTU), method = "jaccard"))) # should be at least 20%
OTUs.left <- as.numeric(gsub("OTU_([0-9]+)", "\\1", rownames(Zheng_OTU)))
Zheng_OTU <- Zheng_OTU[order(OTUs.left), ]
```
```{r, cache=TRUE, eval=FALSE}
write.table(Zheng_OTU, "otuCountTable_13C-CH4_SIP.abundant.txt", sep = "\t", col.names = NA)
try(system("sed -i 's/\"\"/\"Group\"/' otuCountTable_13C-CH4_SIP.abundant.txt", intern = TRUE, ignore.stderr = TRUE))
# try(system("sed -i 's/\"X//g' otuCountTable.abundant.txt", intern = TRUE, ignore.stderr = TRUE))
try(system("sed -i 's/\"//g' otuCountTable_13C-CH4_SIP.abundant.txt", intern = TRUE, ignore.stderr = TRUE))
```

```{r, cache=TRUE}
Taxonomy <- read.table("Zheng_silva.nrv119.taxonomy", header = FALSE, row.names = 1, stringsAsFactors = FALSE)
Taxonomy <- as.data.frame(Taxonomy[which(rownames(Taxonomy) %in% rownames(Zheng_OTU)), ])
rownames(Taxonomy) <- rownames(Zheng_OTU)
# Taxonomy <- as.data.frame(apply(Taxonomy, 2, function(x) gsub("\"","",x))) # remove "
Taxonomy <- data.frame(Size = rowSums(Zheng_OTU), Taxonomy = Taxonomy[, 1])
```
```{r, cache=TRUE, eval=FALSE}
write.table(Taxonomy, "Zheng_silva.nrv119_13C-CH4_SIP.abundant.taxonomy", sep = "\t")
```

### Generate sparcc matrix 
```{r, cache=TRUE, eval=FALSE, engine='bash'}
try(system(SparCC.sh otuCountTable.noSIP.abundant.txt))
```

## Plot networks
### Load matrices
```{r, cache=TRUE}
pvals.mat <- read.table("pvals_two_sided.txt",
                        header = TRUE, sep = "\t")
# remove OTU_id
row.names(pvals.mat) <- pvals.mat[, 1]
pvals.mat <- pvals.mat[, -1]
# set p-values of 0 to a non-zero, small p-value so we can take the logarithm
pvals.mat[pvals.mat == 0] <- 0.000000001
# convert into significance
sig.mat <- -1 * log10(pvals.mat) 
# remove all edges with significance below 1
sig.mat[sig.mat < 1] <- 0
sig.mat <- as.matrix(sig.mat)

cor.mat <- read.table("otuCountTable_13C-CH4_SIP.abundant.txt_sparcc.txt",
                      header = TRUE, sep = "\t")
row.names(cor.mat) <- cor.mat[, 1]
cor.mat <- cor.mat[, -1]
sig.cor.mat <- cor.mat
sig.cor.mat[sig.mat == 0] <- 0
good.pos.cor.mat <- sig.cor.mat
good.pos.cor.mat[sig.cor.mat < 0.3] <- 0

Taxonomy <- read.table("Zheng_silva.nrv119_13C-CH4_SIP.abundant.taxonomy", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
Taxonomy <- PrepOtuTax(Taxonomy, tax.level = "Class")
```

### Plot methanotrophs networks
#### Load methanotroph names
```{r, cache=TRUE}
# load list of methanotrophs
Methanotrophs <- readLines("../Methanotroph_names.txt")

# List only OTUs which match methanotrophs list
to.keep <- numeric(0)
for (i in seq(length(Methanotrophs))) {
  hits <- Taxonomy$Genus == Methanotrophs[i]
  to.keep <- c(to.keep, which(hits == TRUE))
}

Methano.taxonomy <- Taxonomy[to.keep, ]
```

### Make sure only labelled methanotrophs are used
```{r, cache=TRUE}
otuCountTable.abundant <- as.matrix(read.table("otuCountTable_13C-CH4_SIP.abundant.txt", header = TRUE, row.names = 1))# %>% 
  # sweep(., 1, rowSums(.), '/') * 100 
small.samples <- which(colSums(otuCountTable.abundant) < 500)

colnames(otuCountTable.abundant) %>% 
  sub(".*\\.Fraction\\.([0-9]+)\\..*$", "\\1", .) %>%
  as.numeric() %>%
  sapply(., function(x) if (x < 8) {x <- "Heavy"} else {x <- "Light"}) %>%
  as.factor() -> Fractions

# for each methanotroph in the OTU table compare heavy and light fractions between gradients
methano.pvals <- data.frame(pval = rep(0, length(to.keep)), direction = rep(0, length(to.keep)))
gradients <- sub("Day\\.(.*)\\.Fraction.*", "\\1", colnames(otuCountTable.abundant))

for (i in seq(to.keep)) {
  direction <- 1
  p.vect <- 1
  for (n in seq(100)) {
    temp.mat <- otuCountTable.abundant
    temp.mat[1, small.samples] <- 500
    temp.mat <- t(rrarefy(t(temp.mat), 500))
    temp.mat[, small.samples] <- otuCountTable.abundant[, small.samples]

    test.mat <- data.frame(abundance = temp.mat[to.keep[i], ])
    test.mat$gradient <- gradients
    test.mat$fraction <- Fractions
    
    mod <- aov(test.mat$abundance ~ test.mat$fraction + test.mat$gradient)
    direction[n] <- TukeyHSD(mod)[[1]][1]
    p.vect[n] <- TukeyHSD(mod)[[1]][4]
  }
  methano.pvals$direction[i] <- mean(direction)
  methano.pvals$pval[i] <- mean(p.vect)
}

methano.pvals$adj.pval <- p.adjust(methano.pvals$pval, method = "bonferroni")
heavy.methanotrophs <- 0
j <- 1
for (i in seq(nrow(methano.pvals))) {
  if (methano.pvals$adj.pval[i] <= 0.05 & methano.pvals$direction[i] < 0) {
    heavy.methanotrophs[j] <- to.keep[i]
    j <- j + 1
  }
}

Labelled.Methano.taxonomy <- Taxonomy[heavy.methanotrophs, ]
```

#### Calculate methanotrophs network of positive interactions
```{r, cache=TRUE}
# keep only OTU rows interacting with methanos
to.keep <- numeric(0)
for (i in seq(nrow(Labelled.Methano.taxonomy))) {
  hits <- rownames(cor.mat) == rownames(Labelled.Methano.taxonomy)[i]
  to.keep <- c(to.keep, which(hits == TRUE))
}

Methano.pvals.mat <- pvals.mat[to.keep, ]
Methano.cor.mat <- cor.mat[to.keep, ]

# set p-values of 0 to a non-zero, small p-value so we can take the logarithm
Methano.pvals.mat[Methano.pvals.mat == 0] <- 0.000000001
# convert into significance
sig.mat <- -1 * log10(Methano.pvals.mat) 
# remove all edges with significance below 1
sig.mat[sig.mat < 1] <- 0
sig.mat <- as.matrix(sig.mat)

# Extract only OTUs with good correlations
sig.cor.mat <- Methano.cor.mat
sig.cor.mat[sig.mat == 0] <- 0
good.Methano.cor.mat <- Methano.cor.mat
good.Methano.cor.mat[Methano.cor.mat < 0.3 & Methano.cor.mat > -0.3] <- 0
good.Methano.pos.cor.mat <- Methano.cor.mat
good.Methano.pos.cor.mat[Methano.cor.mat < 0.3] <- 0
good.Methano.neg.cor.mat <- Methano.cor.mat
good.Methano.neg.cor.mat[Methano.cor.mat > -0.3] <- 0

### extract remaining columns (otus) plus the methano otus from the original correlation matrix cor.mat
good.Methano.pos.cor.mat <- good.Methano.pos.cor.mat[, colSums(good.Methano.pos.cor.mat) > 0] # this still includes methanotrophs with no associations
# remove methanotrophs with no associations
methano.cols <- colnames(good.Methano.pos.cor.mat) %in% rownames(Labelled.Methano.taxonomy)
methano2remove <- intersect(names(which(colSums(good.Methano.pos.cor.mat[, methano.cols]) == 1)), names(which(rowSums(good.Methano.pos.cor.mat[, ]) == 1)))
# methano2remove <- names(which(colSums(good.Methano.pos.cor.mat[, methano.cols]) == 1))
if (length(methano2remove) > 0) {
  good.Methano.pos.cor.mat <- good.Methano.pos.cor.mat[-which(rownames(good.Methano.pos.cor.mat) %in% methano2remove), -which(colnames(good.Methano.pos.cor.mat) %in% methano2remove)] 
}

OTUs2keep <- unique(c(rownames(good.Methano.pos.cor.mat), colnames(good.Methano.pos.cor.mat))) # actually `colnames(Methano.cor)` is enough because we didn't work with half matrix
indices2keep1 <- numeric(0)
indices2keep2 <- numeric(0)
for (i in seq(length(OTUs2keep))) {
  hits1 <- rownames(good.pos.cor.mat) == OTUs2keep[i]
  indices2keep1 <- c(indices2keep1, which(hits1 == TRUE))
  hits2 <- rownames(Taxonomy) == OTUs2keep[i]
  indices2keep2 <- c(indices2keep2, which(hits2 == TRUE))
}

Relevant.cor.mat <- as.matrix(good.pos.cor.mat[indices2keep1, indices2keep1])
# Relevant.pvals <- as.matrix(pvals[indices2keep, indices2keep])
Relevant.taxa <- Taxonomy[indices2keep2, ]

# mark methanotrophs
to.keep <- numeric(0)
for (i in seq(length(Methanotrophs))) {
  hits <- Relevant.taxa$Genus == Methanotrophs[i]
  to.keep <- c(to.keep, which(hits == TRUE))
}
Relevant.taxa$Methanotrophs <- logical(nrow(Relevant.taxa))
Relevant.taxa$Methanotrophs[to.keep] <- TRUE

# prepare matrix for calculating network
mat4adjm <- Relevant.cor.mat
order.id <- as.numeric(sub("OTU_([0-9]*)", "\\1", rownames(mat4adjm)))
mat4adjm <- mat4adjm[order(order.id), order(order.id)]
Relevant.taxa <- Relevant.taxa[order(order.id), ]

# calc network
sparcc.graph <- graph.adjacency(mat4adjm, mode = c("undirected"), weighted = TRUE, add.rownames = TRUE, diag = FALSE) # generate the network graph object

# add features to graph
sparcc.graph$colour <- as.character(lapply(Relevant.taxa$Class, as.character))
sparcc.graph$abundance <- Relevant.taxa$Size / sum(Relevant.taxa$Size) * 100

sparcc.graph$methanotroph <- Relevant.taxa$Methanotrophs

c(coords, edges) := PrepNetDF(sparcc.graph, layout_function = layout.fruchterman.reingold, membership = TRUE)

(p1 <- GGNetworkPoints(coords, tax.level = "Class", OTU.labels = TRUE)) 
p1 + theme(legend.position = "none")
(p2 <- GGNetworkSegments(edges))
p2 + theme(legend.position = "none") 

(p3 <- GGNetworkMethano(coords, edges, tax.level = "Class", OTU.labels = FALSE))
```
